Including EW and mixed QCD-EW corrections into PDF fits

### 2020-07-27 paper/app-pineappl.tex
Complete section 2 and appendix A

\section{More technical details about \texorpdfstring{\textsc{PineAPPL}}{PineAPPL}}

@@

All parts are written in Rust and therefore need a Rust compiler and other tools.
On computers with a \texttt{bash} shell the installation is as simple as
\begin{lstlisting}[language=bash]
curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh
\end{lstlisting}
which downloads the compiler \texttt{rustc}, the package manager \texttt{cargo}, and a few other helpful tools.
When the installation has completed make sure to read and follow the instructions printed before.
See also \url{https://www.rust-lang.org/tools/install} for more details and for installation instructions for other systems.

@@

The command-line program \texttt{pineappl} is compiled and installed using
\begin{lstlisting}[language=bash]
cargo install pineappl_cli
\end{lstlisting}
This program also needs \texttt{LHAPDF} \cite{Buckley:2014ana} installed.
For usage instructions simply type \texttt{pineappl} in your shell and read the help message.

@@

For the C interface you need to first install \texttt{cargo-c},
\begin{lstlisting}[language=bash]
cargo install cargo-c
\end{lstlisting}
and then download the \textsc{PineAPPL} repository, compile and finally install into it into a directory \texttt{\$prefix} as follows:
\begin{lstlisting}[language=bash]
git clone https://github.com/N3PDF/pineappl/
cd pineappl_capi/
cargo cinstall --release --prefix=${prefix}
\end{lstlisting}
The last line will install the C header \texttt{pineappl\_capi.h}, the library, and a pkg-config\footnote{A standard way on Linux to express how dependencies are compiled/linked against, see \url{https://www.freedesktop.org/wiki/Software/pkg-config/}} file (\texttt{pineappl\_capi.pc}).
Make sure that the environment variables \texttt{PATH}, \texttt{LD\_LIBRARY\_PATH}, and \texttt{PKG\_CONFIG\_PATH} are properly set.
The latter is needed for \texttt{pkg-config -{}-cflags -{}-libs pineappl\_capi} to work, which prints the necessary compiler/linker flags.
See App.~\ref{app:example-program} for an example.

@@

\subsection{Drell--Yan \texorpdfstring{$n$}{n}-tuple storage size}
\label{app:drell-yan-storage}

TODO

@@

\clearpage
\begin{lstlisting}[language=C++,mathescape=true]
// create a new luminosity function for the $\gamma\gamma$ initial state
auto* lumi = pineappl_lumi_new();
int32_t pdg_ids[] = { 22, 22 };
double ckm_factors[] = { 1.0 };
pineappl_lumi_add(lumi, 1, pdg_ids, ckm_factors);

@@

// destroy the object
pineappl_grid_delete(grid);
\end{lstlisting}
\clearpage

@@

TODO

@@

TODO

>>>>

\section{Installation and usage of \texorpdfstring{\textsc{PineAPPL}}{PineAPPL}}

@@

All parts are written in Rust and therefore need a Rust compiler and other tools.
On computers with a \texttt{bash} shell the installation is as simple as
\begin{lstlisting}[language=bash]
 $ curl --proto '=https' --tlsv1.2 -sSf https://sh.rustup.rs | sh
\end{lstlisting}
which downloads the compiler \texttt{rustc}, the package manager \texttt{cargo}, and a few other helpful tools.
When the installation has completed make sure to read and follow the instructions printed before.
See also \url{https://www.rust-lang.org/tools/install} for more details and for installation instructions for other systems.

@@

The command-line program \texttt{pineappl} is compiled and installed using
\begin{lstlisting}[language=bash]
 $ cargo install pineappl_cli
\end{lstlisting}
This program also needs \texttt{LHAPDF} \cite{Buckley:2014ana} installed.
For usage instructions simply type \texttt{pineappl} in your shell and read the help message.

@@

For the C interface you need to first install \texttt{cargo-c},
\begin{lstlisting}[language=bash]
 $ cargo install cargo-c
\end{lstlisting}
and then download the \textsc{PineAPPL} repository, compile and finally install into it into a directory \texttt{\$prefix} as follows:
\begin{lstlisting}[language=bash]
 $ git clone https://github.com/N3PDF/pineappl/
 $ cd pineappl_capi/
 $ cargo cinstall --release --prefix=DIRECTORY
\end{lstlisting}
The last line will install the C header \texttt{pineappl\_capi.h}, the library, and a pkg-config\footnote{A standard way on Linux to express how dependencies are compiled/linked against, see \url{https://www.freedesktop.org/wiki/Software/pkg-config/}} file (\texttt{pineappl\_capi.pc}) into the directory specified as \texttt{DIRECTORY}.
Make sure that the environment variables \texttt{PATH}, \texttt{LD\_LIBRARY\_PATH}, and \texttt{PKG\_CONFIG\_PATH} are properly set.
The latter is needed for \texttt{pkg-config -{}-cflags -{}-libs pineappl\_capi} to work, which prints the necessary compiler/linker flags.

After being installed, you can compile and link against the library.
See App.~\ref{app:example-program} for an example.

@@

@@

\begin{lstlisting}[language=C++,mathescape=true]
// create a new luminosity function for the $\gamma\gamma$ initial state
auto* lumi = pineappl_lumi_new();
int32_t pdg_ids[] = { 22, 22 };
double ckm_factors[] = { 1.0 };
pineappl_lumi_add(lumi, 1, pdg_ids, ckm_factors);

@@

// destroy the object
pineappl_grid_delete(grid);
\end{lstlisting}

@@

The program \texttt{pineappl} can be used to perform quick convolutions and other calculations with existing grids on the command line.
If started without any arguments, it prints its help and lists all supported subcommands:
\begin{lstlisting}[language=bash]
 $ pineappl
pineappl 0.1.0
Read, write, and query PineAPPL grids

USAGE:
    pineappl <SUBCOMMAND>

FLAGS:
    -h, --help       Prints help information
    -V, --version    Prints version information

SUBCOMMANDS:
    channels           Shows the contribution for each partonic channel
    convolute          Convolutes a PineAPPL grid with a PDF set
    help               Prints this message or the help of the given
                       subcommand(s)
    luminosity         Shows the luminosity function
    merge              Merges one or more PineAPPL grids together
    orders             Shows the predictions for all bins for each order
                       separately
    pdf_uncertainty    Calculates PDF uncertainties
\end{lstlisting}
As an example we show the grid produced for the ATLAS DY high-mass lepton-pair production (from section~\ref{sec:atlas-high-mass-dy}).
The output shows all 13 bins with lower (\texttt{xmin}) and upper limit (\texttt{xmax}) of the invariant mass $M_{\ell \bar{\ell}}$ of the lepton pair, with the differential cross section $\mathrm{d} \sigma / \mathrm{d} M_{\ell \bar{\ell}}$ (\texttt{diff}), integrated cross section $(M_{\ell \bar{\ell}}^\mathrm{max} - M_{\ell \bar{\ell}}^\mathrm{min}) \mathrm{d} \sigma / \mathrm{d} M_{\ell \bar{\ell}}$ (\texttt{integ}), together with the perturbative uncertainty estimated from a 7-point scale variation (\texttt{neg unc} and \texttt{pos unc}).
\begin{lstlisting}[language=bash]
 $ pineappl convolute ATLASZHIGHMASS49FB.pineappl \
 > NNPDF31_nlo_as_0118_luxqed
bin xmin xmax     diff        integ     neg unc pos unc
---+----+----+------------+------------+-------+-------
  0  116  130 2.0630698e-1  2.8882978e0  -2.08%   1.69%
  1  130  150 9.1818985e-2  1.8363797e0  -1.79%   1.86%
  2  150  170 4.5306370e-2 9.0612740e-1  -1.60%   1.98%
  3  170  190 2.5894856e-2 5.1789712e-1  -1.66%   2.06%
  4  190  210 1.6075267e-2 3.2150534e-1  -1.70%   2.10%
  5  210  230 1.0526659e-2 2.1053317e-1  -1.71%   2.12%
  6  230  250 7.1928162e-3 1.4385632e-1  -1.71%   2.13%
  7  250  300 4.0776555e-3 2.0388277e-1  -1.70%   2.44%
  8  300  400 1.4775481e-3 1.4775481e-1  -1.94%   2.87%
  9  400  500 4.5473785e-4 4.5473785e-2  -2.30%   3.19%
 10  500  700 1.2164277e-4 2.4328555e-2  -2.41%   3.12%
 11  700 1000 1.9792340e-5 5.9377019e-3  -2.05%   2.12%
 12 1000 1500 2.0228761e-6 1.0114381e-3  -1.29%   0.47%
\end{lstlisting}

@@

The following run card was used to produce the results shown in section~\ref{sec:atlas-high-mass-dy}.
The only difference to a normal \textsc{mg5\_aMC@NLO} run is the switch \texttt{set iappl 1}, which enables filling a \textsc{PineAPPL} grid.
For a complete set of runcards and patches see TODO.

\lstset{
  basicstyle=\ttfamily\small,
  showstringspaces=false,
  commentstyle=\ttfamily\small,
  keywordstyle=\ttfamily\small,
  identifierstyle=\color{black},
  stringstyle=\color{black},
}

\begin{lstlisting}
set complex_mass_scheme True
import model loop_qcd_qed_sm_Gmu
define p = p b b~
define j = p
generate p p > e+ e- [QCD QED]
output @OUTPUT@
launch @OUTPUT@
fixed_order = ON
set mz @MZ@
set ymt @YMT@
set ebeam1 3500
set ebeam2 3500
set pdlabel lhapdf
set lhaid 324900
set fixed_ren_scale True
set fixed_fac_scale True
set mur_ref_fixed @MZ@
set muf_ref_fixed @MZ@
set reweight_scale True
set ptl = 25.0
set etal = 2.5
set mll = 116
#user_defined_cut set mmllmax = 1500.0
set req_acc_FO 0.0001
set iappl 1
done
quit
\end{lstlisting}

### 2020-07-27 paper/sec-doublecounting.tex
Complete section 2 and appendix A

The ability to include theoretical calculations of EW corrections in
PDF determinations, as has been mentioned in the introduction, is not enough to make a \emph{consistent} fit possible. Indeed, experimental data should
be provided in a format that allows PDF collaborations to easily employ them in these fits in a theoretically sound manner. Before concluding the paper,
we will provide some guidelines to facilitate a consistent presentation of experimental data to that end. We focus on the problem of subtracting (partial) EW effects in the data, which, in PDF fits with theory predictions including them, leads to multiple accounting of EW effects (so-called double counting).

>>>>

The ability to include theoretical calculations of EW corrections in
PDF determinations, as has been mentioned in the introduction, is not enough to make a \emph{consistent} fit possible. Indeed, experimental data should
be provided in a format that allows PDF collaborations to easily employ them in these fits in a theoretically sound manner. Before concluding the paper,
we will provide some guidelines to facilitate a consistent presentation of experimental data to that end. We focus on the problem of data with (partially) subtracted EW effects, which, in PDF fits with theory predictions including them, leads to double counting of these EW effects.

### 2020-07-27 paper/sec-pineappl.tex
Complete section 2 and appendix A

In particular, the following features are supported:
\begin{itemize}
\item The inclusion of perturbative corrections (fixed-order, i.e.\ without parton-shower matching) with any given set of powers of $\alpha$, $\alphas$, in particular including mixed QCD--EW corrections.
\item The support for non-coloured initial-state partons, such as photon-initiated contributions.
In fact, \textsc{PineAPPL} allows arbitrary initial-state combinations, e.g.\ leptonic initial states~\cite{Bertone:2015lqa,Buonocore:2020nai}.
\item The estimate of theory uncertainty via variations of the renormalisation and factorisation scale (the electroweak coupling is assumed to be scale-independent, consistently with the commonest renormalisation schemes).
%\item Support for all-order predictions coming from a resummation calculation or a photon-/parton-shower, which are important for some observables (see section~\ref{sec:results}),
\end{itemize}
On a more technical level, we further point out the following differences and improvements in \textsc{PineAPPL}:
\begin{itemize}
\item \textsc{APPLgrid} typically required a substantial amount of memory (close to \SI{120}{\giga\byte} for Drell--Yan), if a grid for a differential distributions with more than 10 bins was generated.
The memory usage is substantially reduced (roughly \SI{1}{\giga\byte}) after \enquote{optimisation} of the grids, i.e.\ an optimisation of the grid representation in memory.
However, this required a two-step procedure: first, to produce an unoptimised grid in order to identify those parts where the cross section is either zero or extremely suppressed; second, after those parts are removed, to fill an optimised grid with a small number of bins.
\textsc{PineAPPL} avoids this by using a more space-efficient representation from the start.
\item \textsc{PineAPPL} offers an easy-to-use interface written in the C programming language, to allow MC integrators to read and write \textsc{PineAPPL} grids.
C was chosen because it can be easily interfaced with both Fortran and C++, in which most MC integrators are written.
The interface consists of roughly 30 functions, among which only a handful are needed in practice.
See appendix~\ref{app:example-program} for an example.
%\textsc{PineAPPL} itself is written in Rust (see appendix~\ref{app:installation} for installation instructions).
\item The shell command \texttt{pineappl}, which performs convolutions on the command line (see appendix~\ref{app:pineappl-demo}), without requiring the user to write a new program.
In addition to convolutions, it can also print how the luminosity function is constructed, which perturbative orders are stored, their size, the size of each partonic channel, etc., separately for each bin.
%\item Possibility to import \textsc{APPLgrids} and \textsc{fastNLO} tables.
\end{itemize}

@@

% TODO: add table with aMCfast notation

%\begin{table}
%\begin{tabular}{ll}
%\toprule
%Old notation & new notation \\
%\midrule
%\bottomrule
%\end{tabular}
%\caption{Translation table for the notation of ref.~\cite{Bertone:2014zva} to the one used in eq.~\eqref{eq:expansion}.}
%\label{tab:amcfast-notation}
%\end{table}

@@

Note that in this section we have omitted the dependence of the weights, observable and scales on the specific kinematics for
which they are computed. Indeed, beyond LO,
different kinematic contributions have to be considered (in ref.~\cite{Bertone:2014zva} they are labelled with an index $\alpha$, see
eq.\ 12 therein). Their consistent treatment must be ensured by the interface between
\textsc{PineAPPL} and the specific cross-section computation at hand\footnote{
    In the FKS subtraction scheme~\cite{Frixione:1995ms,Frixione:1997np} employed in \textsc{mg5\_aMC@NLO}
    maximally one kind of kinematics for each counterterm (soft, collinear, and soft-collinear) is needed, but this
    is not the general case, for example in the Catani--Seymour subtraction~\cite{Catani:1996jh} different dipoles may have different phase spaces and therefore different scales.}.
However, \textsc{PineAPPL} remains completely blind to this fact, as all weights are filled into a given grid
identified only by the indices $k, l, m, n$.

@@

%At NLO this process has 63 different initial states, of which there is one photon--photon initial state and 31 initial states with at least one (anti-)quark.
%The remaining 31 are given by the transposition of the partons from the first 31 initial states.

% TODO: add graphical representation of the multi-coupling expansion

@@

Using 4-tuples has the clear advantage of being very easy to implement and test.
Furthermore, they reproduce the exact numerical value that is also calculated by the MC integrator.
However, the price one has to pay is the size of the 4-tuples, which scales linearly with the number of phase-space weights $N$.
Because the uncertainties of MC integrators shrink only as $1/\sqrt{N}$ this typically means \numrange{e6}{e9} phase-space points are necessary for a good convergence.
In the case of the NLO Drell--Yan (section~\ref{sec:pineappl-example}) the 4-tuples need roughly TODO~\si{\giga\byte} of storage (see appendix~\ref{app:drell-yan-storage} for an explanation of this number).
With increasing size also the speed of the convolution degrades, because it basically becomes bound by the speed with which the 4-tuples can be read from disk.

@@

A different strategy is to partition the $(x_1, x_2, Q^2)$ space,
\begin{equation}
H = [x_\mathrm{min},x_\mathrm{max}]^2 \times [Q^2_\mathrm{min}, Q^2_\mathrm{max}] \ni (x_1, x_2, Q^2)
\end{equation}
along each axis into a small numbers of bins and to insert the phase-space weights $w$ into the corresponding discrete bin.
Using the bin centres and their values one already has a straightforward implementation of eq.~\eqref{eq:weight-map}, but given a finite number of bins it usually yields an insufficient approximation for the cross section.
Increasing the number of bins improves the accuracy, but it also increases the space requirements.
Interpolation methods offer more accuracy with the same number of bins.

@@

This method first maps $(x_1, x_2, Q^2) \mapsto (y_1, y_2, \tau)$, with
\begin{equation}
y(x) = 5 (1-x) - \log x \text{,} \qquad \tau (Q^2) = \log \log \frac{Q^2}{(\SI{0.25}{\giga\electronvolt})^2} \text{.}
\label{eq:maps}
\end{equation}
The function $y(x)$ maps events with large $x$ effectively linearly and small $x$ effectively logarithmically onto $y$.
This reflects our knowledge of PDFs, which behave differently in those regions, and thereby increases the accuracy of the interpolation.
For the convolution of a grid with a PDF set also the inverse functions are needed, which are
\begin{equation}
x(y) = \frac{1}{5} \operatorname{W}_0 (5 \exp (5-y)) \text{,} \qquad Q^2 (\tau) = (\SI{0.25}{\giga\electronvolt})^2 \exp (\exp (\tau)) \text{,}
\end{equation}
where $\operatorname{W}_0 (x)$ is (the principle branch of) the Lambert W function or product logarithm, which satisfies the relation $\operatorname{W} (x) \exp (\operatorname{W} (x)) = x$.

Following ref.~\cite{Carli:2010rw} (see eq.~(17) therein), we divide the weights before filling them into the grid by the function
\begin{equation}
\omega (x_1, x_2) = \left( \frac{\sqrt{x_1}}{1 - 0.99 x_1} \right)^3 \left( \frac{\sqrt{x_2}}{1 - 0.99 x_2} \right)^3 \text{.}
\end{equation}
This flattens the interpolated function in the region $x \to 1$, where the PDFs are small and tend towards zero, and enhances the function in the small-$x$ region.
The effect of this step is an improvement of the accuracy that depends on the initial states and the process, but it can be as large as a factor of \numrange{10}{100} (one or two more correct digits), see section~TODO.
Before performing a convolution this step is inverted by simply multiplying the interpolated grid values with $\omega (x_1, x_2)$.

@@

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we assume we have chosen a static scale, so that we do not need to interpolate $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
To each grid point a numerical value $a_{i,j}$ is associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them\footnote{The technical term for this interpolation problem is sometimes called \emph{subtabulation}.}.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u_1(y_1(x_1)),u_2(y_2(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula:
\begin{equation}
a_{i,j} = a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2)
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting the smallest index in the square, $k_i$ and $k_j$.
We also remapped
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$.

\begin{figure}
\centering
TODO
\caption{TODO}
\label{fig:example-grid}
\end{figure}

Finally, figure~\ref{fig:example-grid} shows an example of a convoluted grid, but differential in $x_1$ and $x_2$.

\subsubsection{Combination of differently weighed phase-spaces points}

To improve the convergence of phase-space integrations MC integrators usually use importance sampling, typically using one or both of VEGAS~\cite{} and multi-channel MC~\cite{}.
Both algorithms are adaptive and perform a number of iterations, each iteration trying to improve the convergence by changing the distribution from which phase-space points are sampled.
This raises the question if the differently sampled phase-space points are compatible with each other in the grid generation, and, if the answer is no, how this can be remedied.

>>>>

In particular, the following features are supported:
\begin{itemize}
\item The inclusion of perturbative corrections (fixed-order, i.e.\ without parton-shower matching) with any given set of powers of $\alpha$, $\alphas$, in particular including mixed QCD--EW corrections.
\item The support for non-coloured initial-state partons, such as photon-initiated contributions.
In fact, \textsc{PineAPPL} allows arbitrary initial-state combinations, e.g.\ leptonic initial states~\cite{Bertone:2015lqa,Buonocore:2020nai}.
\item The estimate of theory uncertainty via variations of the renormalisation and factorisation scale (the electroweak coupling is assumed to be scale-independent, consistently with the commonest renormalisation schemes).
%\item Support for all-order predictions coming from a resummation calculation or a photon-/parton-shower, which are important for some observables (see section~\ref{sec:results}),
\end{itemize}
On a more technical level, we further point out the following differences and improvements in \textsc{PineAPPL}:
\begin{itemize}
\item \textsc{APPLgrid} typically required a substantial amount of memory (close to \SI{120}{\giga\byte} for Drell--Yan), if a grid for a differential distributions with more than 10 bins was generated.
The memory usage is substantially reduced (roughly \SI{1}{\giga\byte}) after \enquote{optimisation} of the grids, i.e.\ an optimisation of the grid representation in memory.
However, this required a two-step procedure: first, to produce an unoptimised grid in order to identify those parts where the cross section is either zero or extremely suppressed; second, after those parts are removed, to fill an optimised grid with a small number of bins.
\textsc{PineAPPL} avoids this by using a more space-efficient representation from the start.
\item \textsc{PineAPPL} offers an easy-to-use interface written in the C programming language, to allow MC integrators to read and write \textsc{PineAPPL} grids.
C was chosen because it can be easily interfaced with both Fortran and C++, in which most MC integrators are written.
The interface consists of roughly 30 functions, among which only a handful are needed in practice.
See appendix~\ref{app:example-program} for an example.
\item The shell command \texttt{pineappl}, which performs convolutions on the command line (see appendix~\ref{app:pineappl-demo}), without requiring the user to write a new program.
In addition to convolutions, it can also print how the luminosity function is constructed, which perturbative orders are stored, their size, the size of each partonic channel, etc., separately for each bin.
%\item Possibility to import \textsc{APPLgrids} and \textsc{fastNLO} tables.
\end{itemize}

@@

@@

Note that in this section we have omitted the dependence of the weights $w$, the observable $\mathcal{O}$, and the scales $\mu_\mathrm{F}, \mu_\mathrm{R}, Q^2$ on the specific kinematics for which they are computed.
Indeed, beyond LO, different kinematic contributions have to be considered (in ref.~\cite{Bertone:2014zva}, for example, they are labelled with an index $\alpha$, see eq.~(12) therein).
In the FKS subtraction scheme~\cite{Frixione:1995ms,Frixione:1997np} employed in \textsc{mg5\_aMC@NLO} maximally one type of kinematics for each counterterm (soft, collinear, and soft-collinear) is needed, but this is not the general case.
In Catani--Seymour subtraction~\cite{Catani:1996jh}, for example, different dipoles have different phase spaces and therefore different scales.
However, \textsc{PineAPPL} remains completely blind to this fact, and consistent treatment is ensured by correctly using the interface between \textsc{PineAPPL}, which fills each event into a grid using the \emph{numerical value} of $Q^2$.

@@

@@

Using 4-tuples has the clear advantage of being very easy to implement and test.
Furthermore, they reproduce the exact numerical value that is also calculated by the MC integrator.
However, the price one has to pay is the size of the 4-tuples.
To give a concrete example, NLO QCD+EW Drell--Yan lepton-pair production (see section~\ref{sec:pineappl-example} and section~\ref{sec:atlas-high-mass-dy}) needs \SI{159}{\giga\byte} of storage for a target precision of \SI{1}{\percent} of the integrated cross section with this precision.
While this is an acceptable size, increasing the precision by an order of magnitude would require roughly 100 times the size, due to the Monte Carlo convergence that goes as $1/\sqrt{N}$ with $N$ being the number of 4-tuples.
With increasing size also the speed of the convolution degrades, because it basically becomes bound by the speed with which the 4-tuples can be read from disk.
However, due to the uncompressed nature of this representation it can serve as a intermediate format to develop and quickly cross check more space-efficient representations, which we will discuss next.

@@

A different strategy is to partition the $(x_1, x_2, Q^2)$ space,
\begin{equation}
H = [x_\mathrm{min},x_\mathrm{max}]^2 \times [Q^2_\mathrm{min}, Q^2_\mathrm{max}] \ni (x_1, x_2, Q^2)
\end{equation}
along each axis into a small numbers of bins and to insert the phase-space weights $w$ into the corresponding discrete bin.
Using the bin centres and their values one already has a straightforward implementation of eq.~\eqref{eq:weight-map}, but given a finite number of bins usually yields an insufficient approximation for the cross section.
Increasing the number of bins improves the precision, but it also increases the space requirements.
Interpolation methods offer more precision using the same number of bins.

@@

This method first maps $(x_1, x_2, Q^2) \mapsto (y_1, y_2, \tau)$, with
\begin{equation}
y(x) = 5 (1-x) - \log x \text{,} \qquad \tau (Q^2) = \log \log \frac{Q^2}{(\SI{0.25}{\giga\electronvolt})^2} \text{.}
\label{eq:maps}
\end{equation}
The function $y(x)$ maps events with large $x$ effectively linearly and small $x$ effectively logarithmically onto $y$.
This reflects our knowledge of PDFs, which behave differently in those regions, and thereby increases the precision of the interpolation.
For the convolution of a grid with a PDF set also the inverse functions are needed, which are
\begin{equation}
x(y) = \frac{1}{5} \operatorname{W}_0 (5 \exp (5-y)) \text{,} \qquad Q^2 (\tau) = (\SI{0.25}{\giga\electronvolt})^2 \exp (\exp (\tau)) \text{,}
\end{equation}
where $\operatorname{W}_0 (x)$ is (the principle branch of) the Lambert W function or product logarithm, which satisfies the relation $\operatorname{W} (x) \exp (\operatorname{W} (x)) = x$.

Following ref.~\cite{Carli:2010rw} (see eq.~(17) therein), we furthermore divide the weights, before filling them into the grid, by the function
\begin{equation}
\omega (x_1, x_2) = \left( \frac{\sqrt{x_1}}{1 - 0.99 x_1} \right)^3 \left( \frac{\sqrt{x_2}}{1 - 0.99 x_2} \right)^3 \text{.}
\end{equation}
This flattens the interpolated function in the region $x \to 1$, where the PDFs are small and tend towards zero, and enhances the function in the small-$x$ region.
The effect of this step is an improvement of the precision that depends on the initial states and the process, but it can be as large as a factor of \numrange{10}{100} (one or two more correct digits compared to the MC result).
Before performing a convolution this step is inverted by simply multiplying the interpolated grid values with $\omega (x_1, x_2)$.

@@

To illustrate the filling step we give an example in figure~\ref{fig:grid}, where, for simplicity, we assume we have chosen a static scale, so that we do not need to interpolate in $\tau$ direction, and where we also limited the number of grid points to $N_y = 10$.
Each grid point has a numerical value $a_{i,j}$ associated, and the set of all numerical values $\{ a_{i,j} \}$ for all grid indices $(i,j) \in [0,N_y) \times [0,N_y)$ constitute \enquote{the grid}.
Inserting a specific weight $W = w (x_1, x_2) / \omega(x_1, x_2)$ into the grid is shown in figure~\ref{fig:grid} as a small black square, inside the larger grey one.
We have defined the grid points at specific positions, but the points given by the MC will land somewhere between them\footnote{The technical term for this interpolation problem is sometimes called \emph{subtabulation}.}.
The interpolation order $s_y$ then defines a square with length $s_y + 1$ around its centre $(u_1(y_1(x_1)),u_2(y_2(x_2)))$, given by the MC.
All grid points with indices $(i,j)$ covered by the grey square are then updated according to the following formula:
\begin{equation}
a_{i,j} = a_{i,j} + I_i(u(y(x_1))) I_j(u(y(x_2))) W(x_1,x_2) / \omega(x_1,x_2)
\label{eq:interpolation}
\end{equation}
with the Lagrange basis functions
\begin{equation}
I_i (u) = \prod_{\substack{k=k_i \\ k \neq i}}^{k_i + s_y} \frac{u-k}{i-k}
\end{equation}
where the product runs over all indices of the grid points covered by the grey square in figure~\ref{fig:grid}, starting the smallest index in the square, $k_i$ and $k_j$.
Finally, we remapped
\begin{equation}
u(y) = \frac{y-y_\text{min}}{\Delta y} \text{,} \quad
\end{equation}
using $y_\text{min} = y(x_\text{max})$ and $y_\text{max} = y(x_\text{min})$ and the grid spacing $\Delta y = (y_\text{max} - y_\text{min})/(N_y-1)$, so that the integer part of $u(y)$ gives the grid index, e.g.\ $u(y_\text{min}) = 0$ and $u(y_\text{max}) = N_y - 1$, and the fractional part gives the location between the nearest grid points.
